---
title: "strings_and_factors"
author: "William Anderson"
date: "2022-10-18"
output: github_document
---


```{r}
library(tidyverse)
library(rvest)
library(p8105.datasets)
```

## Strings and regex

The most frequent operation involving strings is to search for an exact match. You can use str_detect to find cases where the match exists (often useful in conjunction with filter), and you can use str_replace to replace an instance of a match with something else (often useful in conjunction with mutate). In the following examples we’ll mostly focus on vectors to avoid the complication of data frames, but we’ll see those shortly.

Regex = regular expressions

```{r}
string_vec = c("my", "name", "is", "Will")

str_detect(string_vec, "Will")

str_replace(string_vec, "Will", "will")
```

For exact matches, you can designate matches at the beginning or end of a line.

```{r}
string_vec = c(
  "i think we all rule for participating",
  "i think i have been caught",
  "i think this will be quite fun actually",
  "it will be fun, i think")

str_detect(string_vec, "i think$")

```

"^i think" means find things that start with i think

for end of line do i think$, finds lines that end with i think

```{r}
string_vec = c(
  "Y'all remember Pres. HW Bush?",
  "I saw a green bush",
  "BBQ and Bushwalking at Molonglo Gorge",
  "BUSH -- LIVE IN CONCERT!!"
  )

str_detect(string_vec,"[Bb]ush")
```

str_detect(string_vec,"[Bb]ush") is looking for either upper case B in Bush or lower case b in bush but not all caps BUSH

You don’t have to list these; instead, you can provide a range of letters or numbers that count as a match.

```{r}
string_vec = c(
  '7th inning stretch',
  '1st half soon to begin. Texas won the toss.',
  'she is 5 feet 4 inches tall',
  '3AM - cant sleep :('
  )

str_detect(string_vec, "^[0-9][a-zA-Z]")
```

str_detect(string_vec, "^[0-9][a-zA-Z]") this detects a number followed by a lower case or upper case letter at the beginning of string

The character . matches anything.

```{r}
string_vec = c(
  'Its 7:11 in the evening',
  'want to go to 7-11?',
  'my flight is AA711',
  'NetBios: scanning ip 203.167.114.66'
  )

str_detect(string_vec, "7.11")
```

Some characters are “special”. These include [ and ], ( and ), and .. If you want to search for these, you have to indicate they’re special using \. Unfortunately, \ is also special, so things get weird.

```{r}

string_vec = c(
  'The CI is [2, 5]',
  ':-]',
  ':-[',
  'I found the answer on pages [6-7]'
  )

str_detect(string_vec, "\\[")

```

## Why factors are weird

Factors are the way to store categorical variables in R. They can take on specific levels (e.g. male and female) which are usually presented as characters but are, in fact, stored by R as integers. These integer values are used by functions throughout R – in making plots, in organizing tables, in determining the “reference” category – but most of the time are hidden by easier-to-read character string labels. This close relationship to strings, when in fact there is a lot of added structure, is why factors can be so confusing.

This is the kind of thing that can get you in trouble.

```{r}
vec_sex = factor(c("male", "male", "female", "female"))
vec_sex
```

```{r}
as.numeric(vec_sex)
```

```{r}
vec_sex = fct_relevel(vec_sex, "male")
vec_sex
```
level has been reversed to have male as first

```{r}
as.numeric(vec_sex)
```

The previous code also illustrates coersion: forcing a variable from one type (e.g. factor) to another (e.g. numeric). Understanding how R coerces variables is important, because it sometimes happens unintentionally and can break your code or impact your analyses.

For this reason, it’s important to be deliberate factors, particularly by using functions in forcats.

## NSDUH

```{r}
nsduh_url = "http://samhda.s3-us-gov-west-1.amazonaws.com/s3fs-public/field-uploads/2k15StateFiles/NSDUHsaeShortTermCHG2015.htm"

table_marj = 
  read_html(nsduh_url) %>% 
  html_table() %>% 
  first() %>%
  slice(-1)
```

There are a few steps we need to implement to tidy these data. For now I’m not interested in the p-values (I’d rather just see the data); we also have age groups and year ranges in column titles, both of which are, in fact, variables. Lastly, the table includes letters as superscripts next to table entries; if we only want the percents we’ll need to strip these out.

```{r}
data_marj = 
  
  table_marj %>%
  
  select(-contains("P Value")) %>%
  
  pivot_longer(
    -State,
    names_to = "age_year",
    values_to = "percent") %>%
  
  separate(age_year, into = c("age", "year"), sep = "\\(") %>%
  
  mutate(
    year = str_replace(year, "\\)", ""),
    percent = str_replace(percent, "[a-c]$", ""),
    percent = as.numeric(percent)) %>%
  filter(!(State %in% c("Total U.S.", "Northeast", "Midwest", "South", "West")))
```

percent = str_replace(percent, "[a-c]$", "") - this replaces letters a-c at the end of percent values with nothing aka removes them

year = str_replace(year, "\\)", "") - this replaces any parentheses in the year column with nothing aka removes them

filter(!(State %in% c("Total U.S.", "Northeast", "Midwest", "South", "West"))) - this takes out any categories listed in the c() vector

We used stringr and regular expressions a couple of times above:

in separate, we split age and year at the open parentheses using "\\("

we stripped out the close parenthesis in mutate
to remove character superscripts, 

we replaced any character using "[a-c]$"

Let’s quickly visualize these data for the 12-17 age group; to make the plot readable, we’ll treat State as a factor are reorder according to the median percent value.

```{r}
data_marj %>%
  filter(age == "12-17") %>%
  mutate(State = fct_reorder(State, percent)) %>%
  ggplot(aes(x = State, y = percent, color = year)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

mutate(State = fct_reorder(State, percent)) - this orders the state variable in order of increasing percent of marijuana use

## Restaurant inspections

As a kind of involved example of strings and factors, we’ll take a look at the NYC Restuarant Inspections data. Although we won’t talk about this in detail now, it’s worth mentioning that these data were collected using the NYC Open Data API; code is available with the data description.

First we’ll import the data and take a look.

```{r}
data("rest_inspec")

rest_inspec %>%
  group_by(boro, grade) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = grade, values_from = n)
```

To simplify things, I’ll remove inspections with scores other than A, B, or C, and also remove the restaurants with missing boro information. I’ll also clean up boro names a bit.

```{r}
rest_inspec = 
  rest_inspec %>%
  filter(grade %in% c("A", "B", "C"), boro != "Missing") %>%
  mutate(boro = str_to_title(boro))
```

str_to_title(boro)) - converts boro from all caps to just first letter upper case

Let’s focus only on pizza places for now, and re-examine grades by borough.

```{r}
rest_inspec %>%
  filter(str_detect(dba, "Pizza")) %>%
  group_by(boro, grade) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = grade, values_from = n)
```

dba column has the restaurant names but we filtered by strictly "Pizza" while there could be restaurants that are titled "PIZZA" or "pizza"

```{r}
rest_inspec %>%
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) %>%
  group_by(boro, grade) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = grade, values_from = n)
```

now we visualize

```{r}
rest_inspec %>%
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) %>%
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 
```

Might help to have things in a different order – maybe number of pizza places? Fortunately this can be done using fct_infreq.

```{r}
rest_inspec %>%
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) %>%
  mutate(boro = fct_infreq(boro)) %>%
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 
```

Suppose I want to rename a borough. I could try using str_replace.

```{r}
rest_inspec %>% 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) %>%
  mutate(
    boro = fct_infreq(boro),
    boro = str_replace(boro, "Manhattan", "The City")) %>%
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar() 
```

This renamed the borough, but then converted the result back to a string – which, when plotted, was implicitly made a factor and ordered alphabetically. I could switch the order in which I rename and encode the factor order I want, but that might not always work.

If I tried base R (maybe because I found some code online), I might try to do this using replace.

```{r}

rest_inspec %>% 
  filter(str_detect(dba, "[Pp][Ii][Zz][Zz][Aa]")) %>%
  mutate(
    boro = fct_infreq(boro),
    boro = replace(boro, which(boro == "Manhattan"), "The City")) %>%
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar()

```

That didn’t work at all! Factors have very specific values, trying to use a value that is not an existing factor level won’t work.

Fortunately there is a dedicated function for renaming factor levels:

```{r}
rest_inspec %>% 
  filter(str_detect(dba, regex("pizza", ignore_case = TRUE))) %>%
  mutate(
    boro = fct_infreq(boro),
    boro = fct_recode(boro, "The City" = "Manhattan")) %>%
  ggplot(aes(x = boro, fill = grade)) + 
  geom_bar()
```

## Weather data

We saw factors in in Viz Pt 2. In that case, we were interested in reordering a factor variable in order to produce clearer plots – ggplot uses factor levels to determine the order in which categories appear. Let’s revisit those examples now.

Our first step is to load the data we used.

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
```

Our first example reordered name “by hand” using fct_relevel:

```{r}
weather_df %>%
  mutate(name = forcats::fct_relevel(name, c("Waikiki_HA", "CentralPark_NY", "Waterhole_WA"))) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

We could instead reorder name according to tmax values in each name using fct_reorder:

```{r}
weather_df %>%
  mutate(name = forcats::fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

Although you may not have seen linear regression previously, the ordering of factor variables play an important in this case as well. Specifically, the ordering determines the “reference” category, and is something that can be adjusted as needed.

```{r}
weather_df %>%
  lm(tmax ~ name, data = .)
```

```{r}
weather_df %>%
  mutate(name = forcats::fct_relevel(name, c("Waikiki_HA", "CentralPark_NY", "Waterhole_WA"))) %>% 
  lm(tmax ~ name, data = .)
```

In this example, we’re also using a feature of %>% that allows us to specify where the tibble goes: . is a placeholder for the result of the preceding call.

## PULSE data

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  select(id, visit, everything()) %>%
  mutate(
    visit = str_replace(visit, "bl", "00m"),
    visit = factor(visit)) %>%
  arrange(id, visit)

print(pulse_data, n = 12)
```

## Airbnb

We could also use factors in an exploratory analysis of the Airbnb data, for example when looking at the distribution of prices in various neighborhoods. Ordering these according to the median price makes for clearer plots than ordering neighborhoods alphabetically.

```{r}
data("nyc_airbnb")

nyc_airbnb %>%
  filter(neighbourhood_group == "Manhattan") %>% 
  mutate(
    neighbourhood = fct_reorder(neighbourhood, price)) %>% 
  ggplot(aes(x = neighbourhood, y = price)) +
  geom_boxplot() +
  coord_flip() + 
  ylim(0, 1000)
```

